Globally-centered autocovariances in MCMC

### 2020-08-30 Paper/main.tex

\onehalfspacing
\begin{abstract}
Autocovariances are the fundamental quantity of interest in many features of Markov chain Monte Carlo (MCMC) simulations with autcorrelation function (ACF) plots being often used as a visual tool to ascertain the performance of a Markov chain. Unfortunately, for slow mixing Markov chains, the empirical autocovariance can highly underestimate the truth. For multiple chain MCMC sampling,  we propose a globally-centered estimator of autocovariance function (G-ACvF) that pools information from all Markov chains. G-ACvF is also shown to correct the negative bias of single chain empirical autocovariance. We show the impact of this improved estimator in three aspects: (1) ACF plots, (2) estimates of the Monte Carlo asymptotic covariance matrix, and (3) estimates of the effective sample size. We also establish the strong consistency of our improved asymptotic covariance estimator under weak conditions, and obtain its large sample bias and variance. 
\end{abstract}

@@

A key quantity that drives MCMC output analysis is the autocovariance function (ACvF). From their use in autocorrelation (ACF) plots to assessing Monte Carlo variability of estimators, to determining when to stop MCMC simulation, autocovariances drive many visual and quantitative inferences from MCMC output. However, estimators of ACvF are only available for single-chain implementations, with parallel-chain version obtained by naive averaging. As we will demonstrate, this defeats the purpose of running parallel Markov chains from dispersed starting values.

Let $F$ be the target distribution with mean $\mu$ defined on a space $\X \subseteq \mathbb{R}^d$ equipped with a countably generated $\sigma$-field $\mathcal{B}(\X)$. For $s = 1, \dots, m$, let $\{X_{st}\}_{t\geq1}$ be the $s${th} Harris ergodic $F$-invariant Markov chain \citep[see][for definitions]{meyn:twee:2009} employed to learn characteristics about $F$. The process is covariance stationary so that
the ACvF at lag $k$ is defined as 
%
\[
    \Gamma(k) = \Cov_F(X_{s1}, X_{s\,1+k})= \mathbb{E}_F \left[(X_{s1} - \mu)(X_{s\,1+k} - \mu)^T \right]\,.
\]
Estimating $\Gamma(k)$ is critical to assessing the quality of the sampler and the reliability of Monte Carlo estimators. Let $\bar{X}_s = n^{-1} \sum_{t=1}^{n} X_{st}$ denote the Monte Carlo estimator of $\mu$ from the $s$th chain. The standard estimator for $\Gamma(k)$ is the sample autocovariance matrix at lag $k \geq 0$:
%
\begin{equation} \label{eq:empirical_ACvF}
    \hat{\Gamma}_s(k) = \dfrac{1}{n}\sum_{t=1}^{n-k} \left(X_{st} - \bar{X}_s \right) \left(X_{s\, t + k} - \bar{X}_s \right)^T\,,
\end{equation}
and for $k < 0$, $\hat{\Gamma}_s(k) = \hat{\Gamma}_s(-k)^T$. For single-chain MCMC runs, the estimator $\hat{\Gamma}_s(k)$ is used to construct ACF plots, to estimate the long-run variance of Monte Carlo estimators \citep{hannan:1970,dame:1991}, and to estimate effective sample size (ESS) \citep{kass:carlin:gelman:neal:1998,gong:fleg:2016,vats:fleg:jon:2019}. However, there is no unified approach to constructing estimators of $\Gamma(k)$ for parallel-chain implementations. Specifically, parallel MCMC chains are often spread across the state space so as to adequately capture high density areas. For slow mixing chains, the chains take time to traverse the space so that $\bar{X}_s$  are all vastly different. Consequently, $\hat{\Gamma}_s$ typically underestimates $\Gamma$ leading to a false sense of security about the quality of process.

 We propose a globally-centered estimator of ACvF (G-ACvF) that centers the Markov chain around the overall mean vector from all $m$ chains. We show that the bias under stationarity for G-ACvF is lower than $\hat{\Gamma}_s(k)$, and through various examples, empirically demonstrate improved estimation. We employ the G-ACvF estimators to construct ACF plots, which leads to remarkable improvements. A demonstrative example is at the end of this Section.

@@

Estimators of ACvFs are often needed to estimate the long-run variance of Monte Carlo averages. Specifically, spectral variance (SV) estimators  are used to estimate $\Sigma = \sum_{k=-\infty}^{\infty} \Gamma(k)$ \citep{andr:1991,dame:1991,fleg:jone:2010}. We replace $\hat{\Gamma}(k)$ with G-ACvF in SV estimators to obtain globally-centered SV (G-SV) estimator and demonstrate strong consistency under weak conditions. In the spirit of \cite{andr:1991}, we also obtain large-sample bias and variance of the resulting estimator. SV estimators can be prohibitively slow to  compute \citep{liu:fleg:2018}. To relieve the computational burden, we adapt the fast algorithm of \cite{heberle2017fast} for G-SV estimator that dramatically reduces computation time. The G-SV estimator  is employed in the computation of ESS. We will show that using G-SV estimator for estimating $\Sigma$ safeguards users against early termination of the MCMC process.

@@

We demonstrate the striking difference in estimation of $\Gamma(k)$ via ACF plots. Consider a random-walk Metropolis-Hastings sampler for a univariate mixture of Gaussians target density. Let the target density be 
\[
f(x) = 0.7\,f(x; -5, 1) + 0.3\,f(x; 5, 0.5)\,,
\]
where $f(x; a,b)$ is the density of a normal distribution with mean $a$ and variance $b$. 
We set $m = 2$ with starting values distributed to the two modes. The trace plots in Figure~\ref{fig:gaussian-trace} indicates that in the first $10^4$ steps, the chains do not jump modes so that both Markov chains yield significantly different estimates of the population mean, $\mu$. At $10^5$ sample size, both Markov chains have traversed the state space reasonably and yield similar estimates of $\mu$.

In Figure~\ref{fig:gaussian-acf} for $n = 10^4$, we present the ACF plots using the usual locally-centered ACvF and our proposed G-ACF estimator. The blue curve are the respective estimates at $n = 10^5$. At $n = 10^5$ when the chains have similar means, the G-ACF and locally-centered ACF are equivalent.  However, for $n = 10^4$, the locally-centered ACFs critically underestimate the correlation giving user a misleading visual quality of the Markov chain. 
The G-ACF estimator accounts for the discrepancy in sample means between two chains leading to a far improved quality of estimation.

@@

\caption{ACF plots using local-centering (left) and global-centering (right). Histogram estimates at $n = 10^4$ with the blue curve being estimates from $n = 10^5$ for chain 1.}
 % (b) $n = 10^4$; the dotted lines are ACF for each chain and the thick solid line is the ACF averaged over both chains.}
\label{fig:gaussian-acf}
\end{figure}

@@

% Recall that we run $m$ independent $F$-ergodic Markov chains denoted by $\{X_{st}; \; t \in \mathbb{Z}\}$ for the $s^{th}$ chain. 
Let $P$ denote the Markov transition kernel that uniquely determines the ACvF under stationarity. Recall the sample mean of the $s^{th}$ Markov chain is $\bar{X}_s = n^{-1} \sum_{t=1}^{n} X_{st}$ and denote the global mean by $\bar{\bar{X}} = m^{-1}\sum_{s = 1}^{m}\bar{X}_s$. The global mean is a naturally superior estimator of $\mu$ than $\bar{X}_s$. For $k \geq 0$, we define the globally-centered ACvF (G-ACvF) estimator for the $s${th} Markov chain as
%
\begin{equation} \label{eq:G-ACvF}
    \hat{\Gamma}_{G,s}(k) = \dfrac{1}{n} \sum_{t=1}^{n-k}(X_{st}-\bar{\bar{X}})(X_{s(t+k)}-\bar{\bar{X}})^T,
\end{equation}
with $\hat{\Gamma}_{G,s}(k) = \hat{\Gamma}_{G,s}(-k)^T$ for $k < 0$. 
In the event that all $m$ Markov chains have been run long enough that $\bar{X}_s \approx \bar{\bar{X}}$, then  $\hat{\Gamma}_{s} (k) \approx \hat{\Gamma}_{G,s}(k)$. However, for shorter runs or for slow mixing chains, $\Gamma(k)$ is more appropriately estimated by $\hat{\Gamma}_{G,s}$ as it utilizes information from all chains and accounts for disparity between estimates of $\mu$. 
% We quantify the gains in  bias  for the G-ACvF estimator below. 
Let 
\[
\Phi^{(q)} = \sum_{ k= -\infty}^{\infty}\abs{k}^q \Gamma(k)\,,
\]
and let $\Phi^{(1)}$ be denoted by $\Phi$.  The proof of the following theorem is available in Appendix~\ref{appendix:bias}.

{\color{orange} I don't think that the RHS below is correct. In the formulation below, $o(n^{-2})$ is be $o(n^{-1})$}

@@

When $m = 1$, $\hat{\Gamma}_{G,s}(k) = \hat{\Gamma}_{s}(k)
$ the bias results for which can be found in \cite{priestley1981spectral}. Since lag-covariances are typically positive in MCMC, Theorem~\ref{th:G-ACF_bias} implies that the G-ACvF estimators are asymptotically unbiased and exhibit reduced bias in finite samples compared to locally-centered ACvF estimators. The consequences of this are particularly pertinent for the diagonals of $\Gamma(k)$.
 % Special interest is also in the lag-0 autocovariance, that is, the variance of the target distribution. 
% The covariance matrix of the target distribution is the lag-0 covariance and is of particular interest.
%  The bias of the estimator reduces by a factor of $m$:
% \[
% \mathbb{E} \left[\hat{\Gamma}_{G,s}(0) \right] = \Gamma(0) - \dfrac{\Sigma}{mn} - \dfrac{\Phi}{mn^2} + o(n^{-2})
% \]

% A first use of the ACvFs in MCMC is in constructing ACF plots. 
Let the target variance of $i${th} component be $\Gamma^{ii}(k)$. For any component $i$, the autocorrelation is defined as
\[
\rho^{(i)}(k) = \dfrac{\Gamma^{ii}(k)}{\Gamma^{ii}(0)}\,,
\]
and is instrumental in visualizing the serial correlation in components of the Markov chain. The typical estimate of the autocorrelation is constructed from $\hat{\Gamma}_s(k)$. Instead, we advocate for using G-ACvF estimates so that,
\[
\hat{\rho}_{G,s}^{(i)}(k) = \dfrac{ \hat{\Gamma}^{(i)}_{G,s} (k)}{\hat{\Gamma}^{(i)}_{G,s} (0)}.
\]
The globally-centered autocorrelation provides a far more realistic assessment of the correlation structure of marginal components of the chain  as evidenced in Figure~\ref{fig:gaussian-acf}.

@@

\section{Variance estimators for multiple Markov chains} \label{sec:variance_est}

A critical use of autocovariances is in the assessment of Monte Carlo variability of estimators.  Let $g:\X \to \mathbb{R}^p$ be an $F$-integrable function so that interest is in estimating
$\mu_g = \mathbb{E}_F[g(X)]$.
% \[
% \mu_g = \mathbb{E}_F[g(X)] = \int_{\X}g(x)F(dx)\;.
% \]
 % 
Set $\{Y_{st}\}_{t \geq 1} = \{g(X_{st})\}_{t \geq 1}$ for all $s = 1, \dots, m$. Let $\bar{Y}_s = n^{-1}\sum_{t=1}^{n}Y_{st}$  and  $\bar{\bar{Y}} = m^{-1}\sum_{s=1}^{m}\bar{Y}_s$. By Birkhoff's ergodic theorem,  $\bar{\bar{Y}} \to \mu_g$ with probability 1 as $n \to \infty$. An asymptotic sampling distribution may be available via a Markov chain central limit theorem (CLT) if there exists a $p \times p$ positive-definite matrix $\Sigma$ such that
%
\begin{equation}
\label{eq:CLT}
  \sqrt{mn}(\bar{\bar{Y}} - \mu_g) \xrightarrow{d} N(0,\Sigma)\, \quad \text{ where, }
\end{equation}
% where
\begin{equation}
\label{eq:sigma}
  \Sigma = \sum_{k = -\infty}^{\infty} \Cov_F \left( Y_{11}, Y_{1(1+k)} \right) := \sum_{k = -\infty}^{\infty}\Upsilon (k)
\end{equation}
%
The goal in output analysis for MCMC is to estimate $\Sigma$ in order to assess variability in $\bar{\bar{Y}}$ \citep{fleg:hara:jone:2008,roy:2019,vats:rob:fle:jon:2020}. There is a rich literature on estimating $\Sigma$ for single-chain MCMC implementations. The most common are SV estimators \citep{andr:1991,vats:fleg:jon:2018} and batch means estimators \citep{chen:seila:1987,vats:fleg:jon:2019}. Recently, \cite{gupta:vats:2020} constructed a replicated batch means estimator for estimating $\Sigma$ from parallel chains. Batch means estimators are computationally more efficient than SV estimators, whereas SV estimators are more reliable \citep{damerdji:1995,fleg:jone:2010}. Here, we utilize G-ACvF estimators to construct globally-centered SV (G-SV) estimator of $\Sigma$. Using the method of \cite{heberle2017fast}, we also provide a computationally efficient algorithm for obtaining the G-SV estimator.

The locally and globally-centered  estimators of $\Upsilon(k)$ are $\hat{\Upsilon}_s(k)$ and $\hat{\Upsilon}_{G,s}(k)$, respectively, where for $k \geq 0$
\[
\hat{\Upsilon}_s(k) = \dfrac{1}{n} \ds \sum_{t=1}^{n - k} (Y_{st} - \bar{Y}_s)(Y_{s\,t+k} - \bar{Y}_s)^T \quad \text{ and } \quad \hat{\Upsilon}_{G,s}(k) = \dfrac{1}{n} \ds \sum_{t=1}^{n - k} (Y_{st} - \bar{\bar{Y}})(Y_{s\,t+k} - \bar{\bar{Y}})^T\,,
\] 
and let $\hat{\Upsilon}_G(k) = m^{-1} \sum_{k=1}^{m}\hat{\Upsilon}_{G,s}(k)$. SV estimators are constructed as weighted and truncated sums of estimated ACvFs. For some $c \geq 1$, let $w: \mathbb{R} \to [-c,c]$ be a lag window function and $b_n \in \mathbb{N}$ be a bandwidth. The (locally-centered) SV estimator of $\Sigma$ for chain $s$ is
%
%
\begin{equation} \label{eq:sve}
    \hat{\Sigma}_{s} = \sum_{k=-n+1}^{n-1}w\left(\dfrac{k}{b_n}\right)\hat{\Upsilon}_s(k)\,.
\end{equation}
Large-sample properties of $\hat{\Sigma}_s$ have been studied widely. \cite{vats:fleg:jon:2018} provided conditions for strong consistency while \cite{fleg:jone:2010,vats:flegal:2018} provided large-sample bias and variance results. These results also extend naturally to an average SV (ASV) estimator defined as $\hat{\Sigma}_A := m^{-1} \sum_{s=1}^{m}\hat{\Sigma}_{s}$.
% \[
% \hat{\Sigma}_A = \dfrac{1}{m}\sum_{s=1}^{m}\hat{\Sigma}_{s}\;.
% \]

@@

We define the G-SV estimator as the weighted and truncated sum of G-ACvFs
%
\begin{align*}
    \hat{\Sigma}_{G} &= \sum_{k=-n+1}^{n-1}w\left(\dfrac{k}{b_n}\right)\hat{\Upsilon}_{G}(k)\,.
\end{align*}
%
To accurately estimate $\Sigma$, standard conditions are imposed on the lag window, $w$. 
%

\begin{ass}
\label{ass:lag_window}
The lag window $w(x)$ is continuous at $0$ at all but a finite number of points, is a bounded and even function with $w(0)=1$, \; $\int_{-\infty}^{\infty}w^2(x)dx < \infty$, and $\sum_{k = -\infty}^{\infty} \abs{w(k/b_n)} < \infty$.
\end{ass}

Assumption~\ref{ass:lag_window} is standard \citep[see]{ande:1971} and is satisfied by most lag windows. We will use the following popular Bartlett lag window in our simulations:
\[
w(x) = \begin{cases}
1 - \abs{x} & \textrm{for } \abs{x} \leq 1 \\
0 & \textrm{otherwise}\,.
\end{cases}
\]

@@

First, we provide conditions for strong consistency. Strong consistency is particularly important to ensure sequential stopping rules yield correct coverage at termination \citep{glyn:whit:1992}. A critical assumption is that of a strong invariance principle which the following theorem establishes. Let $B(n)$ denotes a standard $p$-dimensional Brownian motion.

\begin{theorem}[\cite{kuel:1976,vats:fleg:jon:2018}]
  \label{thm:kuelbs}
Let $\E_F\|Y_{11}\|^{2+ \delta} < \infty$ for $\delta > 0$ and let $P$ be polynomially ergodic of order $\xi > (q + 1 + \epsilon)/(1 + 2/\delta)$ for $q \geq 1$. Then there exists a $p \times p$ lower triangular matrix $L$ such that $LL^T = \Sigma$, a non-negative function $\psi(n) = n^{1/2 - \lambda}$ for some $\lambda > 0$, a finite random variable $D$, and a sufficiently rich probability space $\Omega$ such that for almost all $\omega \in \Omega$ and for all $n > n_0$,
\[
\left\|\sum_{t=1}^{n}Y_t - n\mu_g - LB(n)\right\| < D\psi(n) \qquad \text{  with probability 1}\,.
\]
\end{theorem}

% \begin{ass}[Strong Invariance Principle (SIP)] \label{ass:sip}
%     We assume the SIP holds for the Markov chains $\{Y_{st}\}_{t \geq 1} \; \forall s$. Here $\{Y_t\}_{t\geq 1}$ is the stochastic process that follows SIP with mean $\mu_g$. Let $\|.\|$ denote the Euclidean norm. Through SIP, there exists a $p \times p$ lower triangular matrix L, an increasing function $\psi$ on the integers, a finite random variable D, and a  sufficiently rich probability space such that, with probability 1, \\
%   $$\left\|\sum_{t=1}^{n}Y_t - n\mu_g - LB(n)\right\| < D\psi(n)$$
  % 
%   Let $\{B(n)\}_{n\geq 0}$ denotes a standard p-dim Brownian motion and $B^{(i)}$ denotes its $i^{th}$ component. As shown in the results \cite{kuelbs1980almost}, SIP will hold for $\psi(n) = n^{1/2 - \lambda, \; \lambda > 0}$ for Markov chains commonly encountered in MCMC settings. We know that $\psi(n)$ is an inherent property of the stochastic process that satisfies the CLT. This implies that $\psi(n)$ is at least $o(\sqrt{n})$. Using Law of Iterative Logarithms (LIL), a tighter bound for $\psi(n)$ is given by \cite{stra:1964} as $\mathcal{O}(\sqrt{n\log \log n})$.
% \end{ass}

@@

Our next two results establish large-sample bias and variance for G-SV and mimic those of $\hat{\Sigma}_s$ which can be found in \cite{hannan:1970}. However, \cite{hannan:1970} proved the results assuming $\mu_g$ was known. Here, we present the results under mild conditions for when $\mu_g$ is replaced by $\bar{\bar{Y}}$. Let $\Sigma^{ij}$ and $\hat{\Sigma}_G^{ij}$ denote the $ij$th element of the matrix $\Sigma$ and $\hat{\Sigma}_G$, repectively.

>>>>

\onehalfspacing
\begin{abstract}
Autocovariances are a fundamental quantity of interest in Markov chain Monte Carlo (MCMC) simulations with autcorrelation function (ACF) plots being often used as a visual tool to ascertain the performance of a Markov chain. Unfortunately, for slow mixing Markov chains, the empirical autocovariance can highly underestimate the truth. For multiple chain MCMC sampling,  we propose a globally-centered estimator of the autocovariance function (G-ACvF) that exhibits theoretical and empirical improvements. We show that the bias of the G-ACvF estimator is smaller than the bias of the current state-of-the-art.  The impact of this improved estimator is evident in three critical output analysis applications: (1) ACF plots, (2) estimates of the Monte Carlo asymptotic covariance matrix, and (3) estimates of the effective sample size. Under weak conditions, we establish strong consistency of an improved asymptotic covariance estimator, and obtain its large sample bias and variance.  
\end{abstract}

@@

A key quantity that drives MCMC output analysis is the autocovariance function (ACvF).
 % From their use in autocorrelation (ACF) plots to assessing Monte Carlo variability of estimators, to determining when to stop MCMC simulation, autocovariances drive many visual and quantitative inferences from MCMC output. However, 
 Estimators of ACvF are only available for single-chain implementations, with a parallel-chain version obtained by naive averaging. As we will demonstrate, this defeats the purpose of running parallel Markov chains from dispersed starting values.

Let $F$ be the target distribution with mean $\mu$ defined on a space $\X \subseteq \mathbb{R}^d$ equipped with a countably generated $\sigma$-field $\mathcal{B}(\X)$. For $s = 1, \dots, m$, let $\{X_{st}\}_{t\geq1}$ be the $s${th} Harris ergodic $F$-invariant Markov chain \citep[see][for definitions]{meyn:twee:2009} employed to learn characteristics about $F$. The process is covariance stationary so that
the ACvF at lag $k$ is defined as 
%
\[
    \Gamma(k) = \Cov_F(X_{s1}, X_{s\,1+k})= \mathbb{E}_F \left[(X_{s1} - \mu)(X_{s\,1+k} - \mu)^T \right]\,.
\]
Estimating $\Gamma(k)$ is critical to assessing the quality of the sampler and the reliability of Monte Carlo estimators. Let $\bar{X}_s = n^{-1} \sum_{t=1}^{n} X_{st}$ denote the Monte Carlo estimator of $\mu$ from the $s$th chain. The standard estimator for $\Gamma(k)$ is the sample autocovariance matrix at lag $k \geq 0$:
%
\begin{equation} \label{eq:empirical_ACvF}
    \hat{\Gamma}_s(k) = \dfrac{1}{n}\sum_{t=1}^{n-k} \left(X_{st} - \bar{X}_s \right) \left(X_{s\, t + k} - \bar{X}_s \right)^T\,,
\end{equation}
and for $k < 0$, $\hat{\Gamma}_s(k) = \hat{\Gamma}_s(-k)^T$. For a single-chain MCMC run, the estimator $\hat{\Gamma}_s(k)$ is used to construct ACF plots, to estimate the long-run variance of Monte Carlo estimators \citep{hannan:1970,dame:1991}, and to estimate effective sample size (ESS) \citep{kass:carlin:gelman:neal:1998,gong:fleg:2016,vats:fleg:jon:2019}. However, there is no unified approach to constructing estimators of $\Gamma(k)$ for parallel-chain implementations. Specifically, parallel chains are often spread across the state space so as to adequately capture high density areas. For slow mixing chains, the chains take time to traverse the space so that $\bar{X}_s$  are all vastly different. Consequently, $\hat{\Gamma}_s(k)$ typically underestimates $\Gamma(k)$ leading to a false sense of security about the quality of process.

 We propose a globally-centered estimator of ACvF (G-ACvF) that centers the Markov chains around the global mean from all $m$ chains. We show that the bias for G-ACvF is lower than $\hat{\Gamma}_s(k)$, and through various examples, demonstrate improved estimation. We employ the G-ACvF estimators to construct ACF plots and a demonstrative example is at the end of this Section.

@@

Estimators of ACvFs are used to estimate the long-run variance of Monte Carlo averages. Specifically, spectral variance (SV) estimators  are used to estimate $\Sigma = \sum_{k=-\infty}^{\infty} \Gamma(k)$ \citep{andr:1991,dame:1991,fleg:jone:2010}. We replace $\hat{\Gamma}_s(k)$ with G-ACvF in SV estimators to obtain globally-centered SV (G-SV) estimator and demonstrate strong consistency under weak conditions. In the spirit of \cite{andr:1991}, we also obtain large-sample bias and variance of the resulting estimator. SV estimators can be prohibitively slow to  compute \citep{liu:fleg:2018}. To relieve the computational burden, we adapt the fast algorithm of \cite{heberle2017fast} for G-SV estimator that dramatically reduces computation time. The G-SV estimator  is employed in the computation of ESS. We will show that using G-SV estimator for estimating $\Sigma$ safeguards users against early termination of the MCMC process.

@@

We demonstrate the striking difference in estimation of $\Gamma(k)$ via ACF plots. Consider a random-walk Metropolis-Hastings sampler for a univariate mixture of Gaussians target density. Let
\[
f(x) = 0.7\,f(x; -5, 1) + 0.3\,f(x; 5, 0.5)\,,
\]
be the target density where $f(x; a,b)$ is the density of a normal distribution with mean $a$ and variance $b$. 
We set $m = 2$ with starting values distributed to the two modes. The trace plots in Figure~\ref{fig:gaussian-trace} indicate that in the first $10^4$ steps, the chains do not jump modes so that both Markov chains yield significantly different estimates of the population mean, $\mu$. At $10^5$ sample size, both Markov chains have traversed the state space reasonably and yield similar estimates of $\mu$.

In Figure~\ref{fig:gaussian-acf} for $n = 10^4$, we present the ACF plots using $\hat{\Gamma}_s(k)$ and our proposed G-ACF estimator. The blue curves are the respective estimates at $n = 10^5$. At $n = 10^5$ when the chains have similar means, the G-ACF and locally-centered ACF are equivalent.  However, for $n = 10^4$, $\hat{\Gamma}_s(k)$ critically underestimate the correlation giving user a misleading visual of the quality of the Markov chain. 
The G-ACF estimator accounts for the discrepancy in sample means between two chains leading to a far improved quality of estimation.

@@

\caption{ACF plots using local-centering (left) and global-centering (right). Histogram estimates at $n = 10^4$ with the blue curve being estimates from $n = 10^5$ for the first chain.}
 % (b) $n = 10^4$; the dotted lines are ACF for each chain and the thick solid line is the ACF averaged over both chains.}
\label{fig:gaussian-acf}
\end{figure}

@@

Let $P$ denote the Markov transition kernel that uniquely determines the ACvF under stationarity. Recall that the sample mean of the $s^{th}$ Markov chain is $\bar{X}_s$ and denote the global mean vector by $\bar{\bar{X}} = m^{-1}\sum_{s = 1}^{m}\bar{X}_s$. The global mean is a naturally superior estimator of $\mu$ compared to $\bar{X}_s$. For $k \geq 0$, we define the globally-centered ACvF (G-ACvF) estimator for the $s${th} Markov chain as
%
\begin{equation} \label{eq:G-ACvF}
    \hat{\Gamma}_{G,s}(k) = \dfrac{1}{n} \sum_{t=1}^{n-k}(X_{st}-\bar{\bar{X}})(X_{s(t+k)}-\bar{\bar{X}})^T,
\end{equation}
with $\hat{\Gamma}_{G,s}(k) = \hat{\Gamma}_{G,s}(-k)^T$ for $k < 0$. In the event that all $m$ Markov chains have been run long enough, $\bar{X}_s \approx \bar{\bar{X}}$, and hence  $\hat{\Gamma}_{s} (k) \approx \hat{\Gamma}_{G,s}(k)$. However, for shorter runs or for slow mixing chains, $\Gamma(k)$ is more appropriately estimated by $\hat{\Gamma}_{G,s}$ as it utilizes information from all chains and accounts for disparity between estimates of $\mu$. 
% We quantify the gains in  bias  for the G-ACvF estimator below. 
Let 
\begin{equation*}
\Phi^{(q)} = \sum_{ k= -\infty}^{\infty}\abs{k}^q \Gamma(k)\,,  
\end{equation*}
%
and let $\Phi^{(1)}$ be denoted by $\Phi$.  The proof of the following theorem is available in Appendix~\ref{appendix:bias}. Let $\|\cdot\|$ denote Euclidean norm.

% {\color{orange} I don't think that the RHS below is correct. In the formulation below, $o(n^{-2})$ is be $o(n^{-1})$}

@@

When $m = 1$, $\hat{\Gamma}_{G,s}(k) = \hat{\Gamma}_{s}(k)
$ the bias results for which can be found in \cite{priestley1981spectral}. Since lag-covariances are typically positive in MCMC, Theorem~\ref{th:G-ACF_bias} implies that the G-ACvF estimators are asymptotically unbiased and exhibit reduced bias in finite samples compared to $\hat{\Gamma}_s(k)$. The consequences of this are particularly pertinent for the diagonals of $\Gamma(k)$.

Let $\Gamma^{ii}(k)$ be the $i${th} component of $\Gamma(k)$. For component $i$, the autocorrelation is defined as
\[
\rho^{(i)}(k) = \dfrac{\Gamma^{ii}(k)}{\Gamma^{ii}(0)}\,,
\]
and is instrumental in visualizing the serial correlation in components of the Markov chain. A standard estimator of the autocorrelation is constructed from $\hat{\Gamma}_s(k)$. Instead, we advocate for using G-ACvF estimates so that,
\begin{equation}
\label{eq:acf}
 \hat{\rho}_{G,s}^{(i)}(k) = \dfrac{ \hat{\Gamma}^{(i)}_{G,s} (k)}{\hat{\Gamma}^{(i)}_{G,s} (0)}. 
\end{equation}
The globally-centered autocorrelation provides a far more realistic assessment of the correlation structure of marginal components of the chain  as evidenced in Figure~\ref{fig:gaussian-acf}.

@@

\section{Long-run variance estimators} \label{sec:variance_est}

A critical use of autocovariances is in the assessment of Monte Carlo variability of estimators.  Let $g:\X \to \mathbb{R}^p$ be an $F$-integrable function so that interest is in estimating
$\mu_g = \mathbb{E}_F[g(X)]$.
% \[
% \mu_g = \mathbb{E}_F[g(X)] = \int_{\X}g(x)F(dx)\;.
% \]
 % 
Set $\{Y_{st}\}_{t \geq 1} = \{g(X_{st})\}_{t \geq 1}$ for all $s = 1, \dots, m$. Let $\bar{Y}_s = n^{-1}\sum_{t=1}^{n}Y_{st}$  and  $\bar{\bar{Y}} = m^{-1}\sum_{s=1}^{m}\bar{Y}_s$. By Birkhoff's ergodic theorem,  $\bar{\bar{Y}} \to \mu_g$ with probability 1 as $n \to \infty$. An asymptotic sampling distribution may be available via a Markov chain central limit theorem (CLT) if there exists a $p \times p$ positive-definite matrix $\Sigma$ such that
%
\begin{equation}
\label{eq:CLT}
  \sqrt{mn}(\bar{\bar{Y}} - \mu_g) \xrightarrow{d} N(0,\Sigma)\, \quad \text{ where, }
\end{equation}
% where
\begin{equation}
\label{eq:sigma}
  \Sigma = \sum_{k = -\infty}^{\infty} \Cov_F \left( Y_{11}, Y_{1(1+k)} \right) := \sum_{k = -\infty}^{\infty}\Upsilon (k)\,.
\end{equation}
%
The goal in output analysis for MCMC is to estimate $\Sigma$ in order to assess variability in $\bar{\bar{Y}}$ \citep{fleg:hara:jone:2008,roy:2019,vats:rob:fle:jon:2020}. There is a rich literature on estimating $\Sigma$ for single-chain MCMC implementations. The most common are SV estimators \citep{andr:1991,vats:fleg:jon:2018} and batch means estimators \citep{chen:seila:1987,vats:fleg:jon:2019}. Recently, \cite{gupta:vats:2020} constructed a replicated batch means estimator for estimating $\Sigma$ from parallel chains. Batch means estimators are computationally more efficient than SV estimators, whereas SV estimators are more reliable \citep{damerdji:1995,fleg:jone:2010}. Here, we utilize G-ACvF estimators to construct globally-centered SV (G-SV) estimator of $\Sigma$. Using the method of \cite{heberle2017fast}, we also provide a computationally efficient algorithm for obtaining the G-SV estimator.

The locally and globally-centered  estimators of $\Upsilon(k)$ are $\hat{\Upsilon}_s(k)$ and $\hat{\Upsilon}_{G,s}(k)$, respectively, where for $k \geq 0$
\[
\hat{\Upsilon}_s(k) = \dfrac{1}{n} \ds \sum_{t=1}^{n - k} (Y_{st} - \bar{Y}_s)(Y_{s\,t+k} - \bar{Y}_s)^T \quad \text{ and } \quad \hat{\Upsilon}_{G,s}(k) = \dfrac{1}{n} \ds \sum_{t=1}^{n - k} (Y_{st} - \bar{\bar{Y}})(Y_{s\,t+k} - \bar{\bar{Y}})^T\,,
\] 
and let $\hat{\Upsilon}_G(k) = m^{-1} \sum_{k=1}^{m}\hat{\Upsilon}_{G,s}(k)$. SV estimators are constructed as weighted and truncated sums of estimated ACvFs. For some $c \geq 1$, let $w: \mathbb{R} \to [-c,c]$ be a lag window function and $b_n \in \mathbb{N}$ be a bandwidth. The (locally-centered) SV estimator of $\Sigma$ for chain $s$ is
%
%
\begin{equation} \label{eq:sve}
    \hat{\Sigma}_{s} = \sum_{k=-n+1}^{n-1}w\left(\dfrac{k}{b_n}\right)\hat{\Upsilon}_s(k)\,.
\end{equation}
Large-sample properties of $\hat{\Sigma}_s$ have been widely studied. \cite{vats:fleg:jon:2018} provide conditions for strong consistency while \cite{fleg:jone:2010,hannan:1970} provide large-sample bias and variance results. These results also extend naturally to an average SV (ASV) estimator defined as $\hat{\Sigma}_A := m^{-1} \sum_{s=1}^{m}\hat{\Sigma}_{s}$.
% \[
% \hat{\Sigma}_A = \dfrac{1}{m}\sum_{s=1}^{m}\hat{\Sigma}_{s}\;.
% \]

@@

We define the G-SV estimator as the weighted and truncated sum of G-ACvFs
%
\begin{align*}
    \hat{\Sigma}_{G} &= \sum_{k=-n+1}^{n-1}w\left(\dfrac{k}{b_n}\right)\hat{\Upsilon}_{G}(k)\,.
\end{align*}
%
Standard conditions are imposed on the lag window, $w$. 
%

\begin{ass}
\label{ass:lag_window}
The lag window $w(x)$ is continuous at all but a finite number of points, is a bounded and even function with $w(0)=1$, \; $\int_{-\infty}^{\infty}w^2(x)dx < \infty$, and $\sum_{k = -\infty}^{\infty} \abs{w(k/b_n)} < \infty$.
\end{ass}

Assumption~\ref{ass:lag_window} is standard \citep[see][]{ande:1971} and is satisfied by most lag windows. We will use the following popular Bartlett lag window in our simulations:
\[
w(x) = \begin{cases}
1 - \abs{x} & \textrm{for } \abs{x} \leq 1 \\
0 & \textrm{otherwise}\,.
\end{cases}
\]

@@

First, we provide conditions for strong consistency. Strong consistency is particularly important to ensure sequential stopping rules in MCMC yield correct coverage at termination \citep{fleg:gong:2015,glyn:whit:1992}. A critical assumption is that of a strong invariance principle which the following theorem establishes. Let $B(n)$ denotes a standard $p$-dimensional Brownian motion.

\begin{theorem}[\cite{kuel:1976,vats:fleg:jon:2018}]
  \label{thm:kuelbs}
Let $\E_F\|Y_{11}\|^{2+ \delta} < \infty$ for $\delta > 0$ and let $P$ be polynomially ergodic of order $\xi > (q + 1 + \epsilon)/(1 + 2/\delta)$ for $q \geq 1$. Then there exists a $p \times p$ lower triangular matrix $L$ with $LL^T = \Sigma$, a non-negative function $\psi(n) = n^{1/2 - \lambda}$ for some $\lambda > 0$, a finite random variable $D$, and a sufficiently rich probability space $\Omega$ such that for all $n > n_0$,
\[
\left\|\sum_{t=1}^{n}Y_t - n\mu_g - LB(n)\right\| < D\psi(n) \qquad \text{  with probability 1}\,.
\]
\end{theorem}

@@

Conditions for strong consistency of $\hat{\Sigma}_s$ for polynomially ergodic Markov chains are in \cite{vats:fleg:jon:2018}. Typically, $b_n = \lfloor n^{\nu} \rfloor$ for some $0< \nu <1$ for which $n^{-1} b_n \log \log n \to 0$, thus Theorem~\ref{th:consistency} presents no added conditions for strong consistency.
